---
title: "STA240 Final Project"
format: pdf
author: "Anthony Zhao, Abby Li, William Yan"
editor: 
  markdown: 
    wrap: 72
---

# Scenario 1

### Customer Arrival

Poisson process (rate = $\lambda$)

-   $T_k$: Arrival time of the $k$th customer

-   $W_k$: Time between the $k-1$th arrival and the $k$th arrival

$$
  W_k = T_k - T_{k-1}.
$$

$W_k \sim Pois(\lambda)$

where $\lambda = 5$ customers per hour

### Service Time

$S_k \sim Exp(\lambda)$

where $\lambda = 6$ customers per hour, so the average customer needs to
wait 1/6 hours = 10 minutes.

## Arrival Times

```{r arrival-times, message=F, warning=F}
library(tidyverse)

# simulating the arrival times of customers throughout the day

# Poisson process (lambda = 5)
# Tk= arrival time of the kth customer
# Wk= time between the k-1th customer arrival and the kth customer arrival where Wk ~ Pois(lambda)

# set parameter
lambdaA <- 5 # in units: customers per hour
opening_time <- hm("10:00")
closing_time <- hm("22:00")
hours <- hour(closing_time) - hour(opening_time) # operating hours: 10am to 10pm
total_time <- hours*60 # operating hours in minutes
lambdaA <- lambdaA/60 # customers per minute
# converting to minutes because our lambda is low, and we can can get greater precision in analyses this way

n <- ceiling(lambdaA*total_time) # max number of customers the store can have throughout the day

# generate W1,..,Wn (calculating the time between the arrival times of 2 customers)
W_sample <- rexp(n, rate= lambdaA)

# calculate T or the arrival times by summing together the Wi arrival times

T_sample <- numeric(n)

for(i in 1:n) {
  T_sample[i] <- sum(W_sample[1:i])
}

# all possible arrival times of customers throughout the day (X minutes after opening)

# however, the store is only open for 12 hours or 720 minutes so we must get rid of the values that fall outside of the range (go beyond 720)

arrival_times <- T_sample[T_sample <= total_time]

arrival_times

opening_time + minutes(floor(max(arrival_times)))

```

### Arrival Times Analysis

In this simulation, the number of customers that will be arriving within
the operating hours is `r length(arrival_times)`, with the first
customer arriving `r floor(min(arrival_times))` minutes after opening
and the last customer arriving
`r ceiling(total_time - max(arrival_times))` minutes before closing

## Serving Times

```{r serving-times}
# given the output from above, simulate the serving times of customers before they leave

# notice that service time is modeled by exp(6)
lambdaS <- 6 # customers per hour
lambdaS <- lambdaS/60 # customers per minute 

# simulate customer's service time 
# n= only simulating the service time for those where T_sample <= total_time
service_times <- rexp(length(arrival_times), rate= lambdaS)

#these are the serving times for each arriving customer before they leave
service_times 

```

### Serving Times Analysis

```{r service-times-graph}
boxplot(service_times, horizontal= TRUE, main= "Service Times", xlab= "Minutes")
```

The average service time is `r round(mean(service_times))` minutes, with
the data skewed right, consistent with an exponential distribution. This
indicates that service times tend to lower.

## Waiting Times

```{r customer-service}

# determining waiting times

# for each observation (customer), calculate when the service begins and when it ends 
# serving ends = service begins + service time
# service begins: either when the customer walks in, or when the previous customer leaves (assumes there is a waiting time)

# compare this to the arrival time 
# if arrival time > time service ends then wait time = 0
# but if arrival time < service time ends then wait time = time service ends- arrival time 

# variable initialization
waiting_times <- numeric(length(arrival_times))  # generating times for each customer
service_start <- numeric(length(arrival_times)) 
service_end <- numeric(length(arrival_times))  
current_end <- numeric(0) # service end time for current customer (i)

# iterate over each customer
for (i in 1:length(arrival_times)) {

  # only includes observations where service time > arrival time => which means there is a wait time
  # gets rid of observations where service < arrival time => 0 wait time
  if (length(current_end) > 0) {
    current_end <- current_end[current_end > arrival_times[i]]
  }
  
 if (length(current_end) == 0) {
   # scenario 1: if there is no waiting time, service starts at the customer arrival
    service_start[i] <- arrival_times[i]
  } else {
    # scenario 2: if there is a waiting time, service starts at the end of the previous customer's end time
    service_start[i] <- min(current_end)
  }
  
  # update the service end time for current customer by adding when service starts and how long the service takes
  service_end[i] <- service_start[i] + service_times[i]
  
  # add this service end time to current end services
  current_end <- c(current_end , service_end[i])
  
  # update waiting time
  waiting_times[i] <- service_start[i] - arrival_times[i]
}


simulation_results <- data.frame(
  customer = 1:length(arrival_times),
  arrival_time = arrival_times,
  service_length = service_times,
  service_start = service_start,
  service_end = service_end,
  waiting_time = waiting_times
)

print(head(simulation_results, 15)) # printing first 15 customers

```

```{r waiting-times-dist}
boxplot(waiting_times, horizontal= TRUE, main= "Waiting Times", xlab= "Minutes")
mean(waiting_times)
```

Waiting times tend to be short, if not zero, and on average, the waiting
time is on average `r round(mean(waiting_times))` minutes.

# Scenario 2

## Arrival and Service

Assumptions:

1.  5 dining tables and L chefs with operating hours 10am - 10pm

2.  each table only seats one customer

3.  service time modeled by an exponential distribution with rate λS =
    3L, so that the more chefs there are, the faster the service times
    become **(this is not very realistic)**

```{r arrival-service-2}
# first, we generate the arrival times similar in scenario 1
lambdaA <- 24 # per hour
opening_time <- hm("10:00")
closing_time <- hm("22:00")
hours <- hour(closing_time) - hour(opening_time)
total_time <- hours*60 # operating hours in minutes
lambdaA <- lambdaA/60 # per minute

n <- ceiling(lambdaA*total_time) # max number of customers
W_sample <- rexp(n, rate= lambdaA)
T_sample <- numeric(n)

for(i in 1:n) {
  T_sample[i] <- sum(W_sample[1:i])
}

arrival_times <- T_sample[T_sample <= total_time]

# next, we generate the service times similar to scenario 1
# make a function to do this
calc_service_times <- function(arrivals, chefs) {
  # Ensure rate is per unit time
  minute_rate = (3*chefs) / 60
  services = rexp(length(arrivals), rate = minute_rate)
  return(services) # in minutes
}
# if we only have one chef
service_times <- calc_service_times(arrivals = arrival_times, chefs = 2)
```

## Waiting Times

To model waiting times, we iterate through the day minute by minute.

```{r waiting-2}

tables <- 5
arrival_times_temp <- arrival_times

# number of people in line each minute
queue_size_history <- numeric(total_time)

# number of tables occupied each minute
occupied_tables_history <- rep(0, total_time)

# timer to track remaining waiting time for each table in the restaurant
# each element is one table in the restaurant
# -1 means empty
# otherwise, number of remaining service minutes
tables_timer <- rep(-1, tables)

# the amount of minutes each customer of that day waited
waiting_times <- numeric(0)

# the arrival_times indices of the people currently in line
# in order to know how long their eventual service time will be
queue <- numeric(0)

# an internal counter separate from the time
customers_entered <- 0
for (i in 1:total_time) {
  occupied_tables_history[i+1] = occupied_tables_history[i]
  
  # update the waiting timer for all occupied tables
  tables_timer[tables_timer > 0] <- tables_timer[tables_timer > 0] - 1
  # update the number of available tables in the next minute
  # based on the number of tables who have finished timers
  occupied_tables_history[i+1] = occupied_tables_history[i+1] - sum(tables_timer == 0)
  # mark the finished tables as available tables for the next minute
  tables_timer[tables_timer == 0] <- tables_timer[tables_timer == 0] - 1
  
  # has the next customer arrived?
  if(length(arrival_times_temp) > 0){
    if(arrival_times_temp[1] < i) {
      # if so, add them to the back of the queue
      queue = c(queue, as.integer(customers_entered+1)) # add 1 for 1-indexing
      # remove the 1st element of arrival_times
      arrival_times_temp = arrival_times_temp[-1]
      # start the waiting timer for this customer by appending 0
      waiting_times = c(waiting_times, 0)
      
      customers_entered = customers_entered + 1
    }
  }
  # are any tables currently open and there is a person in line?
  if(occupied_tables_history[i+1] < tables & length(queue) > 0) {
    # if so, then seat the first person in line
    # at the first available table
    for (j in 1:tables) {
      if(tables_timer[j] == -1) {
        # queue[1] has the customer index of the first person in line
        tables_timer[j] = round(service_times[queue[1]])
        break
      }
    }
    # the next minute there will be one more occupied table
    occupied_tables_history[i+1] = occupied_tables_history[i+1] + 1
    # remove the first person in the queue
    queue = queue[-1]
  }
  # update the waiting time for each person in the queue
  for (customer_index in queue) {
    waiting_times[customer_index] = waiting_times[customer_index] + 1
  }
  # keep track of how long the line is at each minute
  queue_size_history[i] = length(queue)
}

queue_size_history <- queue_size_history[-1]
occupied_tables_history <- occupied_tables_history[-1]

queue_size_history
occupied_tables_history
waiting_times
```
Make plots using these three pieces of data? Tracked across each day
How do the queueing times look?

## Restaurant Profits

Assumptions:

1.  each customer spends \$50 per meal (customers who are still in the queue when the restaurant closes won't pay)

2.  each chef earns a wage of \$40 per hour (paid for the entire duration of the restaurant's operating hours)

### Maximizing Profits
Should we run this simulation multiple times to create a PDF of the total daily profits?
How many chefs should we hire?

## Down-time of Restaurant
How does the occupancy of the restaurant vary throughout the day? Does that inform any of our recommendations?
