---
title: "STA240 Final Project"
format: pdf
author: "Anthony Zhao, Abby Li, William Yan"
editor: 
  markdown: 
    wrap: 72
---

# Scenario 1

### Customer Arrival

Poisson process (rate = $\lambda$)

-   $T_k$: Arrival time of the $k$th customer

-   $W_k$: Time between the $k-1$th arrival and the $k$th arrival

$$
  W_k = T_k - T_{k-1}.
$$

$W_k \sim Pois(\lambda)$

where $\lambda = 5$ customers per hour

### Service Time

$S_k \sim Exp(\lambda)$

where $\lambda = 6$ customers per hour, so the average customer needs to
wait 1/6 hours = 10 minutes.

## Arrival Times

```{r arrival-times, message=F, warning=F}
library(tidyverse)
library(lubridate)

set.seed(121)

# simulating the arrival times of customers throughout the day

# Poisson process (lambda = 5)
# Tk= arrival time of the kth customer
# Wk= time between the k-1th customer arrival and the kth customer arrival where Wk ~ Pois(lambda)

# set parameter
lambdaA <- 5 # in units: customers per hour
opening_time <- hm("10:00")
closing_time <- hm("22:00")
hours <- hour(closing_time) - hour(opening_time) # operating hours: 10am to 10pm
total_time <- hours*60 # operating hours in minutes
lambdaA <- lambdaA/60 # customers per minute
# converting to minutes because our lambda is low, and we can can get greater precision in analyses this way

n <- ceiling(lambdaA*total_time) # max number of customers the store can have throughout the day

# generate W1,..,Wn (calculating the time between the arrival times of 2 customers)
W_sample <- rexp(n, rate= lambdaA)

# calculate T or the arrival times by summing together the Wi arrival times

T_sample <- numeric(n)

for(i in 1:n) {
  T_sample[i] <- sum(W_sample[1:i])
}

# all possible arrival times of customers throughout the day (X minutes after opening)

# however, the store is only open for 12 hours or 720 minutes so we must get rid of the values that fall outside of the range (go beyond 720)

arrival_times_s1 <- T_sample[T_sample <= total_time]

arrival_times_s1

opening_time + minutes(floor(max(arrival_times_s1)))

```

### Arrival Times Analysis

In this simulation, the number of customers that will be arriving within
the operating hours is `r length(arrival_times_s1)`, with the first
customer arriving `r floor(min(arrival_times_s1))` minutes after opening
and the last customer arriving
`r ceiling(total_time - max(arrival_times_s1))` minutes before closing

## Serving Times

```{r serving-times}
# given the output from above, simulate the serving times of customers before they leave

# notice that service time is modeled by exp(6)
lambdaS <- 6 # customers per hour
lambdaS <- lambdaS/60 # customers per minute 

# simulate customer's service time 
# n= only simulating the service time for those where T_sample <= total_time
service_times_s1 <- rexp(length(arrival_times_s1), rate= lambdaS)

#these are the serving times for each arriving customer before they leave
service_times_s1

```


## Time of the day with arrival time

```{r}
# Start time as POSIXct
start_time <- as.POSIXct("10:00", format = "%H:%M", tz = "UTC")

# Add minutes to the start time

time_of_day <- sapply(arrival_times_s1, function(m) {
  m <- round(m)  # Round to nearest whole number
  new_time <- start_time + (m * 60)  # Add minutes converted to seconds
  format(new_time, "%H:%M")  # Format as "HH:MM"
})

print(time_of_day)
```

## Waiting Times

```{r customer-service}

# determining waiting times

# for each observation (customer), calculate when the service begins and when it ends 
# serving ends = service begins + service time
# service begins: either when the customer walks in, or when the previous customer leaves (assumes there is a waiting time)

# compare this to the arrival time 
# if arrival time > time service ends then wait time = 0
# but if arrival time < service time ends then wait time = time service ends- arrival time 

# variable initialization
waiting_times_s1 <- numeric(length(arrival_times_s1))  # generating times for each customer
service_start <- numeric(length(arrival_times_s1)) 
service_end <- numeric(length(arrival_times_s1))  
current_end <- numeric(0) # service end time for current customer (i)

# iterate over each customer
for (i in 1:length(arrival_times_s1)) {

  # only includes observations where service time > arrival time => which means there is a wait time
  # gets rid of observations where service < arrival time => 0 wait time
  if (length(current_end) > 0) {
    current_end <- current_end[current_end > arrival_times_s1[i]]
  }
  
 if (length(current_end) == 0) {
   # scenario 1: if there is no waiting time, service starts at the customer arrival
    service_start[i] <- arrival_times_s1[i]
  } else {
    # scenario 2: if there is a waiting time, service starts at the end of the previous customer's end time
    previous_end <- service_end[i - 1]
    service_start[i] <- max(arrival_times_s1[i], previous_end)
  }
  
  # update the service end time for current customer by adding when service starts and how long the service takes
  service_end[i] <- service_start[i] + service_times_s1[i]
  
  # add this service end time to current end services
  current_end <- c(current_end , service_end[i])
  
  # update waiting time
  waiting_times_s1[i] <- service_start[i] - arrival_times_s1[i]
}


scen1_sim_results <- data.frame(
  customer = 1:length(arrival_times_s1),
  arrival_time = arrival_times_s1,
  service_length = service_times_s1,
  service_start = service_start,
  service_end = service_end,
  waiting_time = waiting_times_s1,
  time_of_day = time_of_day
)

print(head(scen1_sim_results, 5)) # printing first 5 customers

```

### Serving and Waiting Times Analysis

```{r service-times-graph}
# boxplot(service_times, horizontal= TRUE, main= "Service Times", xlab= "Minutes")

scen1_sim_results %>%
  ggplot(aes(x= service_length)) +
  geom_boxplot() +
  labs(
    x= "Minutes",
    title = "Service Times"
  ) +
  theme_minimal()

mean(service_times_s1)

```

The average service time is `r round(mean(service_times_s1))` minutes, with
the data skewed right, consistent with an exponential distribution. This
indicates that service times tend to lower.

```{r waiting-times-dist}
# boxplot(waiting_times_s1, horizontal= TRUE, main= "Waiting Times", xlab= "Minutes")

mean(waiting_times_s1)

scen1_sim_results %>% 
  ggplot(aes(x= waiting_time)) +
  geom_boxplot() +
  labs(
    x= "Minutes",
    title = "Waiting Times"
  ) +
  theme_minimal()

```

Waiting times tends to be slightly right-skewed and on average, the
waiting time is `r round(mean(waiting_times_s1))` minutes.

```{r}

#Label for 30 min interval
breaks <- seq(30, 720, by = 30)
labels <- sprintf("%02d:%02d", 10 + breaks %/% 60, breaks %% 60) 

ggplot(scen1_sim_results, aes(x = time_of_day, y = waiting_times_s1)) +
  geom_point(color = "blue", size = 2, alpha = 0.6) +
    scale_x_discrete(
    breaks = breaks, 
    labels = labels
  ) +
  labs(
    title = "Scatter Plot",
    x = "Time of the Day",
    y = "Waiting Time"
  ) +
  theme_minimal()

```

```{r}

library(ggplot2)


# Custom breaks and labels for 30-minute intervals
breaks <- seq(30, 720, by = 60)
labels <- sprintf("%02d:%02d", 10 + breaks %/% 60, breaks %% 60)

# Scatter plot with x-axis as numeric time in minutes
ggplot(scen1_sim_results, aes(x = arrival_times_s1, y = waiting_times_s1)) +
  geom_point(color = "blue", size = 2, alpha = 0.6) +
  scale_x_continuous(
    breaks = breaks, 
    labels = labels
  ) +
  labs(
    title = "Scatter Plot",
    x = "Time of the Day",
    y = "Waiting Time"
  ) +
  theme_minimal()+
  theme(panel.grid.minor = element_blank())

```

# Scenario 2

## Arrival and Service

Assumptions:

1.  5 dining tables and L chefs with operating hours 10am - 10pm

2.  each table only seats one customer

3.  service time modeled by an exponential distribution with rate λS =
    3L, so that the more chefs there are, the faster the service times
    become **(this is not very realistic)**

```{r arrival-service-2}
# first, we generate the arrival times similar in scenario 1
lambdaA <- 24 # per hour
opening_time <- hm("10:00")
closing_time <- hm("22:00")
hours <- hour(closing_time) - hour(opening_time)
total_time <- hours*60 # operating hours in minutes
lambdaA <- lambdaA/60 # per minute
num_chefs = 2

n <- ceiling(lambdaA*total_time) # max number of customers
W_sample <- rexp(n, rate= lambdaA)
T_sample <- numeric(n)

for(i in 1:n) {
  T_sample[i] <- sum(W_sample[1:i])
}

arrival_times <- T_sample[T_sample <= total_time]

# next, we generate the service times similar to scenario 1
# make a function to do this
calc_service_times <- function(arrivals, chefs) {
  # Ensure rate is per unit time
  minute_rate = (3*chefs) / 60
  services = rexp(length(arrivals), rate = minute_rate)
  return(services) # in minutes
}
# if we only have one chef
service_times <- calc_service_times(arrivals = arrival_times, chefs = num_chefs)
```

## Waiting Times

To model waiting times, we iterate through the day minute by minute.

```{r waiting-2}

tables <- 5
arrival_times_temp <- arrival_times

# number of people in line each minute
queue_size_history <- numeric(total_time)

# number of tables occupied each minute
occupied_tables_history <- rep(0, total_time)

# timer to track remaining waiting time for each table in the restaurant
# each element is one table in the restaurant
# -1 means empty
# otherwise, number of remaining service minutes
tables_timer <- rep(-1, tables)

# the amount of minutes each customer of that day waited
waiting_times <- numeric(0)

# the arrival_times indices of the people currently in line
# in order to know how long their eventual service time will be
queue <- numeric(0)

# an internal counter separate from the time
customers_entered <- 0
for (i in 1:total_time) {
  occupied_tables_history[i+1] = occupied_tables_history[i]
  
  # update the waiting timer for all occupied tables
  tables_timer[tables_timer > 0] <- tables_timer[tables_timer > 0] - 1
  # update the number of available tables in the next minute
  # based on the number of tables who have finished timers
  occupied_tables_history[i+1] = occupied_tables_history[i+1] - sum(tables_timer == 0)
  # mark the finished tables as available tables for the next minute
  tables_timer[tables_timer == 0] <- tables_timer[tables_timer == 0] - 1
  
  # has the next customer arrived?
  if(length(arrival_times_temp) > 0){
    if(arrival_times_temp[1] < i) {
      # if so, add them to the back of the queue
      queue = c(queue, as.integer(customers_entered+1)) # add 1 for 1-indexing
      # remove the 1st element of arrival_times
      arrival_times_temp = arrival_times_temp[-1]
      # start the waiting timer for this customer by appending 0
      waiting_times = c(waiting_times, 0)
      
      customers_entered = customers_entered + 1
    }
  }
  # are any tables currently open and there is a person in line?
  if(occupied_tables_history[i+1] < tables & length(queue) > 0) {
    # if so, then seat the first person in line
    # at the first available table
    for (j in 1:tables) {
      if(tables_timer[j] == -1) {
        # queue[1] has the customer index of the first person in line
        tables_timer[j] = round(service_times[queue[1]])
        break
      }
    }
    # the next minute there will be one more occupied table
    occupied_tables_history[i+1] = occupied_tables_history[i+1] + 1
    # remove the first person in the queue
    queue = queue[-1]
  }
  # update the waiting time for each person in the queue
  for (customer_index in queue) {
    waiting_times[customer_index] = waiting_times[customer_index] + 1
  }
  # keep track of how long the line is at each minute
  queue_size_history[i] = length(queue)
}

occupied_tables_history <- occupied_tables_history[-1]

```

```{r}
scen2_sim_results_by_customer <- data.frame(
  customer = 1:length(arrival_times),
  arrival_time = arrival_times,
  service_length = service_times,
  waiting_time = waiting_times
)

```

```{r}
scen2_sim_results_by_minute <- data.frame(
  minutes_since_opening = 1:total_time,
  time_of_day = I(lapply(1:total_time, function(i) opening_time + minutes(i))),
  queue_size = queue_size_history,
  occupied_tables = occupied_tables_history
)

```


```{r scenario-2-plots}


scen2_sim_results_by_customer |>
  ggplot(aes(x = arrival_time)) +
  geom_histogram(bins = 12, color = "black", fill = "white") +
  labs(
    x = "Arrival Time (Minutes after Opening)",
    y = "Count"
  ) +
  theme_minimal()

scen2_sim_results_by_customer |>
  ggplot(aes(x = service_length)) +
  geom_histogram(bins = 12, color = "black", fill = "white") +
  labs(
    x = "Service Duration (Minutes)",
    y = "Count"
  ) +
  theme_minimal()

scen2_sim_results_by_customer |>
  ggplot(aes(x = waiting_time)) +
  geom_histogram(bins = 12, color = "black", fill = "white") +
  labs(
    x = "Waiting Time (Minutes)",
    y = "Count"
  ) +
  theme_minimal()

scen2_sim_results_by_customer |>
  ggplot(aes(x = arrival_time, y = waiting_time)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_x_continuous(breaks = seq(0, total_time, by = 60)) +
  labs(
    x = "Minutes Since Opening",
    y = "Waiting time of Customer"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())


scen2_sim_results_by_minute |>
  ggplot(aes(x = minutes_since_opening, y = queue_size)) +
  geom_line() +
  scale_y_continuous(breaks = seq(1, max(queue_size_history), by = 1)) +
  scale_x_continuous(breaks = seq(0, total_time, by = 60)) +
  labs(
    x = "Minutes Since Opening",
    y = "Queue Length"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

scen2_sim_results_by_minute |>
  ggplot(aes(x = minutes_since_opening, y = occupied_tables)) +
  geom_area(fill = "lightblue") +
  geom_line() +
  scale_y_continuous(breaks = seq(1, tables, by = 1)) +
  scale_x_continuous(breaks = seq(0, total_time, by = 60)) +
  labs(
    x = "Minutes Since Opening",
    y = "Number of Tables Occupied"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

scen2_sim_results_by_minute |>
  ggplot(aes(x = minutes_since_opening, y = occupied_tables)) +
  geom_area(fill = "lightblue") +
  geom_line() +
  scale_y_continuous(breaks = seq(1, tables, by = 1)) +
  scale_x_continuous(breaks = seq(0, total_time, by = 60)) +
  labs(
    x = "Minutes Since Opening",
    y = "Number of Tables Occupied"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())


```

```{r}
View(scen2_sim_results_by_minute)
```

## Restaurant Profits

Assumptions:

1.  each customer spends \$50 per meal (customers who are still in the
    queue when the restaurant closes won't pay)

2.  each chef earns a wage of \$40 per hour (paid for the entire
    duration of the restaurant's operating hours)

3.  Customer will not wait longer than 30 minutes

```{r profit-chefs}

# assumption of five tables and 3 chefs

# setting number of chefs to a certain number
# chefs <- 3
# meal <- 50 #50 dollars per meal
# wage <- 12*40 #40 dollars per hour for 12 hrs of work
# revenue <- total_customers * meal
# costs <- wage*chefs
# profit <- revenue - costs
# profit

# only including customers who are waiting for less than 30 mins because will leave the line if wait time is greater than that amount 

filtered_customers <- which(waiting_times < 30) 
total_customers <- length(filtered_customers)
total_customers

meal <- 50 #50 dollars per meal
wage <- 12*40 #40 dollars per hour for 12 hrs of work

profit_table <- data.frame(
  Chefs = integer(),
  Revenue = numeric(),
  Costs = numeric(),
  Profit = numeric()
) 

# if people get mad, the meal will only be 25 because they are mad

for (chefs in 1:10) {

  revenue <- total_customers * meal

  # Calculate costs (wage per chef * number of chefs)
  costs <- wage * chefs

  # Calculate profit
  profit <- revenue - costs

  profit_table <- rbind(profit_table, data.frame(Chefs = chefs, Revenue = revenue, Costs = costs, Profit = profit))

}

print(profit_table)

```

```{r profit-chefs-graph}

```

### Maximizing Profits

Should we run this simulation multiple times to create a PDF of the
total daily profits? How many chefs should we hire?

## Down-time of Restaurant

How does the occupancy of the restaurant vary throughout the day? Does
that inform any of our recommendations?
